Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2020-11-06 manuscript.rmd
1. Corrections while reading

- Deep-water access is arguably the most long-standing, but under-studied mechanisms that plants employ to survive during drought. Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves. We validated ERD estimates against existing empirical data of potential water access depths.
- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 1981-2015. Species with deeper ERD spent did not cross critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
\
Keywords: rooting depths, deep-water access, hydraulic vulnerability and safety margins, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts

@@

\
Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation, stem water storage, deep-water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008]. Deep-water access is arguably the most long-standing, yet under-studied mechanism. As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to those in soil pore water at different depths. However, such data are rare [@Evaristo:2016]. DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

\
  Characterising the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019; @Lu.2020] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydrological droughts (rather than meteorological droughts) realized through trees’ water access--the interaction between rooting depths and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
\
  Here, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation. We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period that experienced droughts of a variety of intensity, frequency and duration. Our hypotheses (see Table \@ref(tab:hypo)) are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water; and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich tropical forest.

@@

To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model. We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We define ERD as the depth at which soil water potential best explained species-specific growth dynamics. The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species). We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations. We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD. We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity. Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent beyond its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.

@@

\
  ELM-FATES simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes). Conservation of these terms is expressed as follows:

\begin{equation}
\begin{split}
\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},i} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
\end{split}
(\#eq:water-balance)
\end{equation}
where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr). The soil profile is discretized into 15 exponentially distributed soil layers by depth $z$ (user-specified range), where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21} m.

@@

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (@Stein.1987) of this global parameter space. In the Conrad catchment soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted GWC to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\mathrm{\theta}_{\textrm{soil}} = \Psi_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\mathrm{\Psi}_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\
We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station over 1985-2018 initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode (see https://biogeodb.stri.si.edu/physical_monitoring/research/barrocolorado for meteorological data and @Faybishenko.2018 for QA/QC procedures). As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) local stream discharge ($q_\textrm{run}$) ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: (1) a long-term (2012-2018) record of average VWC (vertical time domain reflectometry (TDR) probes from three locations near the eddy flux tower over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$)), and (2) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 (depths of 0.15, 0.4 and 1 m, 1299 samples covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).

@@

based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s. $W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.

@@

We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration. We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near the vertical probes (see [SI](#SIData.VWC)).

@@

To estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq. \@ref(eq:M-chosen) and those in SI). Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions). We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant, and thus to determine or co-determine whole plant hydraulic response to soil drying; notably, root hydraulic vulnerability may also be a strong driver, and recent studies suggest that its vulnerability would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein). $K_\textrm{leaf}$ in our models, being based on vulnerability curves, is dynamic and thus depends on rooting depth and rhizosphere water potential. The use of maximum $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019]. To account for the non-linear impact of VPD on growth, we used predicted GPP (hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI). Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.

\
 For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z}
(\#eq:M-chosen)
\end{equation}
where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term. See SI for equations for all the alternate models tested.

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
 Checking through each $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth. Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big(z\,\Big|\, \textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|z,h}| r_{s|z,h} \ge 0.71\Big)\Big)
(\#eq:erd)
\end{equation}
where $L$ is the likelihood of $z$ given the observed growth time series, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.

@@

In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI) on growth. We assumed species-specific mean seasonal curves for LAI (normalized between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime (@Osnas.2018; see SI).

>>>>

- Deep-water access is arguably the most long-standing, but under-studied mechanisms that plants employ to survive during drought. Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves. We validated ERD estimates against existing empirical data of potential water access depths.
- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 35 years (1981-2015). Species with deeper ERD did not cross critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
- The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
\
Keywords: rooting depths, deep-water access, hydraulic vulnerability and safety margins, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts

@@

\
Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation, stem water storage, deep-water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008]. Deep-water access is arguably the most long-standing, yet under-studied mechanism. As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant-available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to those in soil pore water at different depths. However, such data are rare [@Evaristo:2016]. DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required. In general, community-scale data collection for rooting or water-sourcing depths is a formidable challenge in species-rich tropical forests.

\
  Characterizing the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019; @Lu.2020] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydrological (rather than meteorological) droughts realized through trees’ water access--the interaction between rooting depths and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.

\
  Here, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation. We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period that experienced droughts of a variety of intensity, frequency and duration. Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Table \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich tropical forest.

@@

To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model. We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We define ERD as the depth at which soil water potential best explained species-specific growth dynamics. The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species). We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations. We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to seven species that overlapped with our assessment of ERD. We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity. Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent beyond its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.

@@

\
  ELM-FATES simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes). Conservation of these terms is expressed as follows:

\begin{equation}
\begin{split}
\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
\end{split}
(\#eq:water-balance)
\end{equation}
where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr). The soil profile is discretized into maximum 15 exponentially distributed soil layers with layer node depth $z$. Here, $z$ $\in$ $Z$; $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13} m.

@@

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space. In the Conrad catchment soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\mathrm{\theta}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\
We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station over 1985-2018 initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode (see https://biogeodb.stri.si.edu/physical_monitoring/research/barrocolorado for meteorological data and @Faybishenko.2018 for QA/QC procedures). As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous [@Condit:2000] and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (1) evapotranspiration ET from the eddy flux tower by the 50 ha plot ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\cong$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across vertical (0-15 cm depth) time domain reflectometry (TDR) probes from three locations near the eddy flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).

@@

based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s. $W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.

@@

We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration. We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a location near the vertical probes (see [SI](#SIData.VWC)).

@@

To inversely estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq. \@ref(eq:M-chosen) and those in SI). Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions). We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant, and thus to determine or co-determine whole plant hydraulic response to soil drying; notably, root hydraulic vulnerability may also be a strong driver, and recent studies suggest that its vulnerability would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein). $K_\textrm{leaf}$ in our models, being based on vulnerability curves, is dynamic and thus depends on rooting depth and rhizosphere water potential. The use of maximum diurnal $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019]. To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see SI). Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.

\
 For the best model (see model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z,h}
(\#eq:M-chosen)
\end{equation}
where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1 ( $K_{\textrm{leaf},s,i|z,h}$ standardised within species), and $\epsilon$ is the model error term. See SI for equations for all the alternate models tested.

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
 Checking through each $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth. Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of this valid set of depths, which can be described as :

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big(z\,\Big|\, \textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|z,h}| r_{s|z,h} \ge 0.71\Big)\Big)
(\#eq:erd)
\end{equation}
where $L$ is the likelihood of $z$ given the observed growth time series $G_{t}$, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 (=0.4) m as sufficient resolution for model validation.

@@

In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI) on growth. We assumed species-specific mean seasonal curves for LAI (standardised between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime (@Osnas.2018; see SI).
